---
title: "Madhume_Evidence_HW1 "
author: "Evidence Madhume"
output:
  
  pdf_document: default
  html_document: default
  date: "Fall A 2024"
  word_document: default
---
<!-- name of chunk = set_global_options and include chunk in doc -->
```{r set_global_options, echo=TRUE}
 knitr::opts_chunk$set(echo = TRUE)
```


```{r echo=TRUE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)


cbc <- as.data.frame(read_excel("C:\\Users\\actua\\OneDrive\\SMU - MSBA 2024\\Programming for Analytics - ITOM 6253\\Assignment 1\\Charles_BookClub_modified.xls"))
cbc <- cbc[,1:18] 
 str(cbc)
summary(cbc)

```
#### Count the unique values of each variable.
```{r}
cbc_counts <- cbc %>% summarise_all(~ n_distinct(.))
cbc_counts
```

#### Convert Gender and Florence to Factors
```{r}
cbc$Gender <- factor(cbc$Gender, levels = c(0, 1), labels =  c("Female", "Male"))

cbc$Florence <- factor(cbc$Florence, levels = c(0, 1), labels =  c("No", "Yes"))
```

#### Identify the numeric columns for which outlier detection is desired
```{r}
# Check if the columns M and R exist in the data frame
if (!("M" %in% names(cbc)) || !("R" %in% names(cbc))) {
  stop("Columns M and R do not exist in the data frame.")
}

# Check the types of M and R
is_numeric_M <- is.numeric(cbc$M)
is_numeric_R <- is.numeric(cbc$R)

# Print results
cat("Is column M numeric? ", is_numeric_M, "\n")
cat("Is column R numeric? ", is_numeric_R, "\n")

# If both are numeric, you can proceed with outlier detection
if (is_numeric_M && is_numeric_R) {
  cat("Both columns M and R are numeric. Proceed with outlier detection.\n")
} else {
  cat("One or both columns are not numeric. Please check the data types.\n")
}

outvars <- c("M", "R", "F", "FirstPurch")

```


#### Find outliers and set them to missing
Note the use of the *anonymous* function in the following code:
```{r}
# Checking for non-numeric values in columns M and R
cbc$M <- as.numeric(cbc$M)  
cbc$R <- as.numeric(cbc$R)  
cbc$M[cbc$M <= 0] <- NA  

# Printing of NA values 
cat("Number of NA values in M:", sum(is.na(cbc$M)), "\n")
cat("Number of NA values in R:", sum(is.na(cbc$R)), "\n")




```

#### Summary also counts the number of missing values
```{r}

summary(cbc)

```
#### Identify variables for which imputation of missing values is desired
```{r}
# find it in all data frames 
missvars <- sapply(cbc, function(x) sum(is.na(x)))

# (the values are missing in the columns)
vars_with_na <- names(missvars[missvars > 0])

# Printing of columns that need imputation  
cat("Columns with missing values that require imputation:\n")
print(vars_with_na)


```

#### Impute missing values of columns with missing values
```{r}

cbc$M[is.na(cbc$M)] <- median(cbc$M, na.rm = TRUE)  
cbc$R[is.na(cbc$R)] <- median(cbc$R, na.rm = TRUE)  

summary(cbc)


```

#### Delete rows for which there are no books purchased.
```{r}
cbc_no_purchase <- cbc[rowSums(cbc[8: 14] > 0) > 0, ]
summary(cbc_no_purchase)
nrow(cbc)
nrow(cbc_no_purchase)

```

#### Sum the purchases of each book type.
```{r}
cbc_sums <- cbc %>% summarise(across(c(8:17), sum))
cbc_sums
```

Produce a bar chart for the counts of each of the ten books purchased

```{r}
library(psych)
multi.hist(cbc[, 4:7], nrow=2, ncol=2, global = FALSE)
```

Produce a bar chart for the counts of each of the ten books purchased

```{r}
cbc_pivot <- pivot_longer(cbc_sums, cols=c( 1:10))
names(cbc_pivot) <- c("Type", "Sum")

ggplot(data=cbc_pivot, aes(y=Sum, x=Type)) +
  geom_bar(stat = 'identity')
```


### A custom function for calculating 4 moments

```{r}
library(e1071)
#browser()
calcfourstats <- function(x) {
  mu <- round(mean(x, na.rm = TRUE), 2)
  sigma <- round(sd(x, na.rm = TRUE), 2)
  skew <- round(skewness(x, na.rm = TRUE), 3)
  kurt <- round(kurtosis(x, na.rm = TRUE), 2)
  result <- data.frame(mu, sigma, skew, kurt)
  return(result)
}

# Col names and corresponding variable names
columnNames <- c("M", "R", "F", "FirstPurch")
varList <- c("Monetary Value", "Last Purchase", "Frequency", "First Purchase")

# Initialize results data frame
results <- data.frame()

# moments for each column
for (col in columnNames) {
  results <- rbind(results, calcfourstats(cbc_no_purchase[[col]]))
}

# Setting of row names for the results
rownames(results) <- varList

# Print results
print(results)

```

### Creating RFM factors
#### Calculate HML cutoffs for RFM
```{r}
cbc_rfm <- data.frame(lapply(cbc_no_purchase[c("R", "F", "M")], 
  function(x) {
    quantile(x, probs = c(0.33, 0.67, 1.00), na.rm = TRUE) }))

rownames(cbc_rfm) <- c("Low Quantile", "Medium Quantile", "High Quantile")

#Printing the percentiles
print(cbc_rfm)
```

Verify results and test subsetting    
```{r}
cbc_rfm

# Calculate the 33rd percentile of the "M" column and add it to the data frame
cbc_rfm["33%", "M"] <- quantile(cbc_no_purchase$M, probs = 0.33, na.rm = TRUE)
 
cbc_rfm["33%", "M"] #What is the 33rd percentile of M?
```

Create three new variables for HML quantiles of RFM variables
```{r}
cbcRFM <- cbc_no_purchase %>%
  mutate(
    rRFM = if_else(R <= cbc_rfm["Low Quantile", "R"], "L",
                   if_else(R >= cbc_rfm["High Quantile", "R"], "H", "M")),
    
    fRFM = if_else(F <= cbc_rfm["Low Quantile", "F"], "L",
                   if_else(F >= cbc_rfm["High Quantile", "F"], "H", "M")),
    
    mRFM = if_else(M <= cbc_rfm["Low Quantile", "M"], "L",
                   if_else(M >= cbc_rfm["High Quantile", "M"], "H", "M"))
  )

# View the updated dataset with RFM factors
head(cbcRFM)

```
Convert the new HML variables into ordered factors
```{r}
# Convert the new HML variables into ordered factors
cbcRFM[c("rRFM", "fRFM", "mRFM")] <- data.frame(lapply(cbcRFM[c("rRFM", "fRFM", "mRFM")], 
  function(x) {
    factor(x, levels = c("L", "M", "H"), ordered = TRUE)
  }))

# View the updated dataset
head(cbcRFM)
str(cbcRFM)

# Summarize the data
sumTable <- cbcRFM %>% 
  group_by(rRFM, fRFM, mRFM) %>%
  summarize(meanM = round(mean(M, na.rm = TRUE), 2), .groups = 'drop')

# Display the summary table
print(sumTable)

```

#### Make three tables, one for each level of factor mRFM

```{r, echo=TRUE, message=FALSE, warning=FALSE}
for (i in c("L", "M", "H")) {
  shortTable <- xtabs(meanM ~ rRFM + fRFM, sumTable %>% filter(mRFM == i)) 
  print(paste('Monetary Value Segment =', i))
  print(shortTable)
  cat("\n") 
}

```


Median monetary value per visit by gender

```{r}
visitValue <- cbcRFM %>%
  group_by(factor(Gender, labels = c("Female", "Male"))) %>%
  summarize(medianM = round(median(M / F), 2))

visitValue
```

Bubble plot of M by R by Gender

```{r}
ggplot(cbcRFM, aes(x = R, y = M, col = factor(Gender, labels = c("Female", "Male")), size = FirstPurch)) +
  geom_point(alpha = .20) +
  
  labs(x = "Recency", y = "Monetary Value") +
  facet_wrap(~ factor(Gender, labels = c("Female", "Male")), labeller = label_parsed) +
  theme(legend.position = "bottom", legend.box = "vertical", 
        legend.key = element_rect(colour = 'white', fill = 'white'))
```




